---
title: "Introduction to ACSets"
author: "Simon Frost"
date: "`r Sys.Date()`"
output: rmarkdown::html_vignette
vignette: >
  %\VignetteIndexEntry{Introduction to ACSets}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
---

```{r setup, include = FALSE}
knitr::opts_chunk$set(
  collapse = TRUE,
  comment = "#>"
)
```

## Introduction

**Attributed C-Sets** (ACSets) are a category-theoretic generalization of
relational databases and graphs. If you're familiar with data frames or SQL
tables, you can think of an ACSet as a collection of interlinked tables where
the links (foreign keys) and data columns are all specified up front by a
**schema**.

More formally, an ACSet is a *functor* from a schema category to the category
of sets — but you don't need to know category theory to use this package. The
key ideas are:

- **Objects** are like tables (e.g., vertices, edges).
- **Morphisms** (homs) are like foreign keys linking one table to another
  (e.g., every edge has a source vertex and a target vertex).
- **Attributes** attach data values to objects
  (e.g., a vertex label or an edge weight).

This framework is powerful enough to represent directed graphs, Petri nets,
wiring diagrams, and many other structured data types — all with a uniform API
for creation, querying, mutation, and serialization.

The `acsets` package is an R port of the
[AlgebraicJulia ACSets.jl](https://github.com/AlgebraicJulia/ACSets.jl) package.

```{r}
library(acsets)
```

## Schemas

A schema defines the *shape* of your data. Let's start with the simplest
interesting example: a directed graph.

A directed graph has two kinds of things:

- **V** — vertices
- **E** — edges

And two relationships:

- **src** — every edge has a source vertex
- **tgt** — every edge has a target vertex

```{r}
SchGraph <- BasicSchema(
  obs       = c("V", "E"),
  homs      = list(hom("src", "E", "V"),
                   hom("tgt", "E", "V")),
  attrtypes = character(),
  attrs     = list()
)
SchGraph
```

The `hom()` helper creates a morphism specification. Its three arguments are
the morphism name, the domain (source object), and the codomain (target
object). Here, `hom("src", "E", "V")` means "src is a function from edges to
vertices".

You can inspect a schema with accessor functions:

```{r}
objects(SchGraph)
homs(SchGraph)
```

### Adding attributes

Plain graphs are useful, but often we want to attach data — labels, weights,
colours. Attributes serve this purpose. Let's define a **labelled weighted
graph**:

```{r}
SchLWGraph <- BasicSchema(
  obs       = c("V", "E"),
  homs      = list(hom("src", "E", "V"),
                   hom("tgt", "E", "V")),
  attrtypes = c("String", "Numeric"),
  attrs     = list(attr_spec("label",  "V", "String"),
                   attr_spec("weight", "E", "Numeric"))
)
```

The `attr_spec()` helper works like `hom()` but the codomain is an *attribute
type* rather than an object. Attribute types are symbolic names (like
`"String"` or `"Numeric"`) — they tell the schema what kind of data lives in
that column, but the actual R type is determined at runtime.

```{r}
attrs(SchLWGraph)
attrtypes(SchLWGraph)
```

The `arrows()` function returns all arrows (both morphisms and attributes),
and `dom()` / `codom()` look up the domain and codomain of any arrow by name:

```{r}
dom(SchLWGraph, "src")
codom(SchLWGraph, "weight")
```

## Creating and Populating ACSets

### From a schema

Given a schema, you can create an empty ACSet:
```{r}
g <- ACSet(SchGraph)
g
```

### Adding parts

Use `add_part()` to add a single part (row) to an object (table). It returns
the new part's integer ID. Use `add_parts()` to add multiple parts at once.

```{r}
v1 <- add_part(g, "V")
v2 <- add_part(g, "V")
v3 <- add_part(g, "V")
c(v1, v2, v3)
```

When adding edges, you can set morphism values inline:

```{r}
e1 <- add_part(g, "E", src = 1, tgt = 2)
e2 <- add_part(g, "E", src = 2, tgt = 3)
e3 <- add_part(g, "E", src = 3, tgt = 1)
```

Batch addition with `add_parts()`:

```{r}
g2 <- ACSet(SchGraph)
invisible(add_parts(g2, "V", 4))
invisible(add_parts(g2, "E", 3, src = c(1, 2, 3), tgt = c(2, 3, 4)))
nparts(g2, "V")
nparts(g2, "E")
```

### The factory pattern

For repeated use of the same schema, `acset_type()` creates a dedicated
constructor function. The `index` argument specifies which morphisms or
attributes should be indexed for fast reverse lookups.

```{r}
Graph <- acset_type(SchGraph, name = "Graph", index = c("src", "tgt"))
```

The constructor accepts initial data — object counts and arrow values:

```{r}
g <- Graph(V = 4, E = 4,
           src = c(1, 1, 2, 3),
           tgt = c(2, 3, 3, 4))
g
```

This is the most convenient way to create ACSets. Let's also create a
constructor for our labelled weighted graph schema:

```{r}
LWGraph <- acset_type(SchLWGraph, name = "LWGraph", index = c("src", "tgt"))
lg <- LWGraph(V = 3, E = 2,
              src = c(1, 2), tgt = c(2, 3),
              label = c("A", "B", "C"),
              weight = c(1.5, 2.5))
```

## Querying ACSets

### subpart: reading values

`subpart()` is the primary read function. It retrieves morphism or attribute
values for one or more parts.

```{r}
# Source vertex of edge 1
subpart(g, 1, "src")

# Target vertex of edge 2
subpart(g, 2, "tgt")
```

Pass `NULL` to get values for *all* parts:

```{r}
subpart(g, NULL, "src")
subpart(g, NULL, "tgt")
```

For the labelled graph, reading attributes works the same way:

```{r}
subpart(lg, NULL, "label")
subpart(lg, 1, "weight")
```

### Composed subparts

You can follow a chain of morphisms by passing a character vector. For
example, to get the *label* of the *source vertex* of edge 1:

```{r}
subpart(lg, 1, c("src", "label"))
```

This is equivalent to `subpart(lg, subpart(lg, 1, "src"), "label")` but more
concise.

### incident: reverse lookup

`incident()` finds all parts whose subpart value matches a given value. For
example, "which edges have vertex 2 as their target?"

```{r}
incident(g, 2, "tgt")
```

With indexing enabled (as we did with `index = c("src", "tgt")`), this is an
O(1) lookup.

```{r}
# All edges originating from vertex 1
incident(g, 1, "src")
```

### Data frames and tables

`as_data_frame()` converts a single object to a data frame:

```{r}
as_data_frame(g, "E")
```

`tables()` returns all objects as a named list of data frames:

```{r}
tables(lg)
```

### Utility queries

```{r}
nparts(g, "V")
parts(g, "E")
has_part(g, "V", 3)
has_part(g, "V", 99)
has_subpart(g, "src")
has_subpart(g, "colour")
```

### Query DSL

The package provides a SQL-like query interface using `From()`, `Where()`, and
`Select()`, which can be chained with the pipe operator:

```{r}
# Find all edges from vertex 1
From(g, "E") |>
  Where("src", `==`, 1) |>
  Select("src", "tgt")
```

`Where()` accepts any binary comparison operator:

```{r}
# Edges with weight > 2
From(lg, "E") |>
  Where("weight", `>`, 2) |>
  Select("src", "tgt", "weight")
```

## Mutation

**ACSets are mutable.** They use reference semantics (backed by R
environments), so modifying an ACSet changes it in place — copies are *not*
made automatically. Use `copy_acset()` when you need an independent copy.

### set_subpart

Set a single morphism or attribute value:

```{r}
g_mut <- Graph(V = 3, E = 2, src = c(1, 2), tgt = c(2, 3))
as_data_frame(g_mut, "E")

set_subpart(g_mut, 1, "tgt", 3)
as_data_frame(g_mut, "E")
```

### set_subparts

Set multiple morphisms/attributes at once for a given part:

```{r}
set_subparts(g_mut, 2, src = 3, tgt = 1)
as_data_frame(g_mut, "E")
```

### clear_subpart

Reset a value to `NA`:

```{r}
wg <- LWGraph(V = 2, E = 1, src = 1, tgt = 2, weight = 10)
subpart(wg, 1, "weight")

clear_subpart(wg, 1, "weight")
subpart(wg, 1, "weight")
```

### Reference semantics

Because ACSets are mutable, assignment does *not* create an independent copy:

```{r}
g_ref <- g_mut
set_subpart(g_ref, 1, "src", 3)
# g_mut is also changed!
subpart(g_mut, 1, "src")
```

Use `copy_acset()` for a true deep copy:

```{r}
g_copy <- copy_acset(g_mut)
set_subpart(g_copy, 1, "src", 1)
# g_mut is unaffected
subpart(g_mut, 1, "src")
subpart(g_copy, 1, "src")
```

## Deletion

### rem_part: pop-and-swap

`rem_part()` removes a part using a **pop-and-swap** strategy: the last part
takes the deleted part's ID. This gives O(1) deletion but means part IDs can
change. Morphisms pointing to the removed part are cleared to `NA`.

```{r}
g_del <- Graph(V = 4, E = 3, src = c(1, 2, 3), tgt = c(2, 3, 4))
as_data_frame(g_del, "E")

# Remove vertex 2: vertex 4 (the last) takes ID 2
rem_part(g_del, "V", 2)
as_data_frame(g_del, "V")
```

Edges that referenced the removed vertex have their values cleared, while
edges that referenced the last vertex (4) are updated to the new ID (2):

```{r}
as_data_frame(g_del, "E")
```

### rem_parts: batch removal

```{r}
g_del2 <- Graph(V = 5, E = 0)
rem_parts(g_del2, "V", c(2, 4))
nparts(g_del2, "V")
```

### cascading_rem_part: cascading deletion

`cascading_rem_part()` removes a part **and** all parts in other objects that
reference it via morphisms. This is analogous to `ON DELETE CASCADE` in SQL.

```{r}
g_casc <- Graph(V = 4, E = 4,
                src = c(1, 1, 2, 3),
                tgt = c(2, 3, 3, 4))
cat("Before: V =", nparts(g_casc, "V"), ", E =", nparts(g_casc, "E"), "\n")
as_data_frame(g_casc, "E")

# Remove vertex 3: all edges that reference vertex 3 are also removed
cascading_rem_part(g_casc, "V", 3)
cat("After:  V =", nparts(g_casc, "V"), ", E =", nparts(g_casc, "E"), "\n")
as_data_frame(g_casc, "E")
```

## JSON Serialization

ACSets can be serialized to and from JSON, making it easy to exchange data
with other ACSet implementations (Python, Julia, TypeScript).

### In-memory round-trip

```{r}
g_json <- Graph(V = 3, E = 3,
                src = c(1, 2, 3),
                tgt = c(2, 3, 1))

json_data <- generate_json_acset(g_json)
str(json_data, max.level = 2)
```

Parse it back using the same constructor:

```{r}
g_parsed <- parse_json_acset(Graph, json_data)
acset_equal(g_json, g_parsed)
```

### File round-trip

```{r}
tmp <- tempfile(fileext = ".json")
write_json_acset(g_json, tmp)

g_from_file <- read_json_acset(Graph, tmp)
acset_equal(g_json, g_from_file)
```

```{r include = FALSE}
unlink(tmp)
```

You can also serialize the schema itself:

```{r}
str(generate_json_schema(SchGraph))
```

## Example: Petri Net

A **Petri net** is a bipartite graph used to model concurrent systems. It has:

- **S** — species (places)
- **T** — transitions
- **I** — input arcs (from a species to a transition)
- **O** — output arcs (from a transition to a species)

Let's model an **SIR epidemiological model** as a Petri net.

### Define the schema

```{r}
SchPetriNet <- BasicSchema(
  obs       = c("S", "T", "I", "O"),
  homs      = list(hom("is", "I", "S"),
                   hom("it", "I", "T"),
                   hom("os", "O", "S"),
                   hom("ot", "O", "T")),
  attrtypes = c("Name"),
  attrs     = list(attr_spec("sname", "S", "Name"),
                   attr_spec("tname", "T", "Name"))
)
```

### Build the SIR model

```{r}
PetriNet <- acset_type(SchPetriNet, name = "PetriNet",
                       index = c("is", "it", "os", "ot"))
sir <- PetriNet()

# Species
s_S <- add_part(sir, "S", sname = "S")
s_I <- add_part(sir, "S", sname = "I")
s_R <- add_part(sir, "S", sname = "R")

# Transitions
t_inf <- add_part(sir, "T", tname = "infection")
t_rec <- add_part(sir, "T", tname = "recovery")

# Infection: S + I → 2I
invisible(add_part(sir, "I", is = s_S, it = t_inf))
invisible(add_part(sir, "I", is = s_I, it = t_inf))
invisible(add_part(sir, "O", os = s_I, ot = t_inf))
invisible(add_part(sir, "O", os = s_I, ot = t_inf))

# Recovery: I → R
invisible(add_part(sir, "I", is = s_I, it = t_rec))
invisible(add_part(sir, "O", os = s_R, ot = t_rec))
```

### Inspect the model

```{r}
tables(sir)
```

### Query the model

Which species feed into the infection transition?

```{r}
inf_inputs <- incident(sir, t_inf, "it")
inf_inputs
```

What are their names? We can use composed subparts:

```{r}
subpart(sir, inf_inputs, c("is", "sname"))
```

Which species are produced by recovery?

```{r}
rec_outputs <- incident(sir, t_rec, "ot")
subpart(sir, rec_outputs, c("os", "sname"))
```

### Cascading deletion

What happens if we remove the infection transition?

```{r}
sir2 <- copy_acset(sir)
cascading_rem_part(sir2, "T", t_inf)
```

All input and output arcs connected to infection are removed:

```{r}
tables(sir2)
```

## Example: Social Network

Let's model a social network with people and friendships.

### Define the schema

```{r}
SchSocial <- BasicSchema(
  obs       = c("Person", "Friendship"),
  homs      = list(hom("person1", "Friendship", "Person"),
                   hom("person2", "Friendship", "Person")),
  attrtypes = c("String"),
  attrs     = list(attr_spec("name",  "Person",     "String"),
                   attr_spec("since", "Friendship",  "String"))
)

SocialNet <- acset_type(SchSocial, name = "SocialNet",
                        index = c("person1", "person2"))
```

### Build the network

```{r}
net <- SocialNet()

alice <- add_part(net, "Person", name = "Alice")
bob   <- add_part(net, "Person", name = "Bob")
carol <- add_part(net, "Person", name = "Carol")
dave  <- add_part(net, "Person", name = "Dave")

invisible(add_part(net, "Friendship",
                   person1 = alice, person2 = bob,   since = "2020-01-15"))
invisible(add_part(net, "Friendship",
                   person1 = alice, person2 = carol, since = "2021-06-01"))
invisible(add_part(net, "Friendship",
                   person1 = bob,   person2 = dave,  since = "2022-03-10"))
invisible(add_part(net, "Friendship",
                   person1 = carol, person2 = dave,  since = "2023-09-20"))
```

```{r}
tables(net)
```

### Query: Who are Alice's friends?

```{r}
# Friendships where Alice is person1
alice_out <- incident(net, alice, "person1")
subpart(net, alice_out, c("person2", "name"))
```

### Query with the DSL

```{r}
From(net, "Friendship") |>
  Where("person1", `==`, alice) |>
  Select("person1", "person2", "since")
```

### Disjoint union

Suppose we have two separate networks and want to merge them:

```{r}
net1 <- SocialNet()
invisible(add_parts(net1, "Person", 2, name = c("Xander", "Yara")))
invisible(add_part(net1, "Friendship",
                   person1 = 1, person2 = 2, since = "2024-01-01"))

net2 <- SocialNet()
invisible(add_parts(net2, "Person", 2, name = c("Zoe", "Will")))
invisible(add_part(net2, "Friendship",
                   person1 = 1, person2 = 2, since = "2024-06-15"))

merged <- disjoint_union(net1, net2)
tables(merged)
```

Note how the IDs from `net2` are offset in the merged result — the friendship
from `net2` now links persons 3 and 4 instead of 1 and 2.

### Structural equality

```{r}
net_copy <- copy_acset(net)
acset_equal(net, net_copy)
```
