---
title: "Advanced ACSet Patterns"
author: "Simon Frost"
date: "`r Sys.Date()`"
output: rmarkdown::html_vignette
vignette: >
  %\VignetteIndexEntry{Advanced ACSet Patterns}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
---

```{r setup, include = FALSE}
knitr::opts_chunk$set(
  collapse = TRUE,
  comment = "#>"
)
```

```{r}
library(acsets)
```

This vignette covers advanced usage patterns for the `acsets` package. For an
introduction to schemas, ACSets, and basic operations, see
`vignette("acsets")`.

## Custom Schemas

The introduction showed graphs and Petri nets. Here we build schemas for three
other domains to illustrate the flexibility of the ACSet framework.

### Food Web

A food web records which species eat which others. The **Predation** object is
a join table linking a predator species to a prey species — a classic
many-to-many relationship.

```{r}
SchFoodWeb <- BasicSchema(
  obs       = c("Species", "Predation"),
  homs      = list(hom("predator", "Predation", "Species"),
                   hom("prey",     "Predation", "Species")),
  attrtypes = c("String", "Numeric"),
  attrs     = list(attr_spec("name",      "Species",   "String"),
                   attr_spec("trophic",   "Species",   "Numeric"),
                   attr_spec("rate",      "Predation", "Numeric"))
)
FoodWeb <- acset_type(SchFoodWeb, name = "FoodWeb",
                      index = c("predator", "prey"))
```

```{r}
fw <- FoodWeb(
  Species   = 4,
  Predation = 4,
  name      = c("Grass", "Rabbit", "Fox", "Eagle"),
  trophic   = c(1, 2, 3, 3),
  predator  = c(2, 3, 3, 4),
  prey      = c(1, 2, 1, 2),
  rate      = c(0.5, 0.3, 0.1, 0.2)
)
tables(fw)
```

What does the fox eat? We find predation links where fox is predator, then
follow the `prey` morphism to get species names:

```{r}
fox <- 3
fox_links <- incident(fw, fox, "predator")
subpart(fw, fox_links, c("prey", "name"))
```

### State Machine

A finite state machine has states and labelled transitions. Here the
**event** morphism is a self-referential join through an intermediary
**Transition** object.

```{r}
SchFSM <- BasicSchema(
  obs       = c("State", "Transition"),
  homs      = list(hom("source", "Transition", "State"),
                   hom("target", "Transition", "State")),
  attrtypes = c("String"),
  attrs     = list(attr_spec("state_name", "State",      "String"),
                   attr_spec("event",      "Transition", "String"))
)
FSM <- acset_type(SchFSM, name = "FSM", index = c("source", "target"))
```

A traffic light controller:

```{r}
light <- FSM(
  State      = 3,
  Transition = 3,
  state_name = c("Red", "Green", "Yellow"),
  source     = c(1, 2, 3),
  target     = c(2, 3, 1),
  event      = c("timer", "timer", "timer")
)
tables(light)
```

Trace a path: starting from "Red", follow two transitions:

```{r}
current <- 1
for (step in 1:3) {
  trans <- incident(light, current, "source")
  evt <- subpart(light, trans[1], "event")
  nxt <- subpart(light, trans[1], "target")
  cat(subpart(light, current, "state_name"), "--[", evt, "]-->",
      subpart(light, nxt, "state_name"), "\n")
  current <- nxt
}
```

### Database-Style Schema: University

ACSets naturally model relational databases. Here is a three-table schema for
students, courses, and enrolments.

```{r}
SchUniv <- BasicSchema(
  obs       = c("Student", "Course", "Enrolment"),
  homs      = list(hom("student", "Enrolment", "Student"),
                   hom("course",  "Enrolment", "Course")),
  attrtypes = c("String", "Numeric"),
  attrs     = list(attr_spec("sname",  "Student",   "String"),
                   attr_spec("cname",  "Course",    "String"),
                   attr_spec("credits","Course",    "Numeric"),
                   attr_spec("grade",  "Enrolment", "String"))
)
Univ <- acset_type(SchUniv, name = "Univ",
                   index = c("student", "course"))
```

```{r}
db <- Univ(
  Student   = 3,
  Course    = 2,
  Enrolment = 4,
  sname     = c("Alice", "Bob", "Carol"),
  cname     = c("Linear Algebra", "Category Theory"),
  credits   = c(3, 4),
  student   = c(1, 1, 2, 3),
  course    = c(1, 2, 2, 1),
  grade     = c("A", "B+", "A-", "B")
)
tables(db)
```

Which students are enrolled in Category Theory (course 2)?

```{r}
ct_enrol <- incident(db, 2, "course")
subpart(db, ct_enrol, c("student", "sname"))
```

## Indexing and Performance

When you create an ACSet with `index = c("src", "tgt")`, every call to
`incident()` on those arrows uses an O(1) hash-table lookup instead of an O(n)
linear scan. This makes a dramatic difference for large datasets.

### When to Index

Index an arrow when you frequently call `incident()` on it. Common cases:

- **Foreign keys you query in reverse** — e.g., "which edges point to vertex
  *v*?" requires an index on `tgt`.
- **Cascading deletion** — `cascading_rem_part()` internally calls `incident()`
  for every morphism targeting the deleted object's type. Indexing those
  morphisms speeds up deletion.

Do **not** index arrows you only read in the forward direction with
`subpart()`, since `subpart()` is always O(1) regardless of indexing.

### Benchmarking

```{r}
SchBench <- BasicSchema(
  obs       = c("V", "E"),
  homs      = list(hom("src", "E", "V"), hom("tgt", "E", "V")),
  attrtypes = character(),
  attrs     = list()
)
GraphIdx    <- acset_type(SchBench, index = c("src", "tgt"))
GraphNoIdx  <- acset_type(SchBench, index = character())

n_v <- 100
n_e <- 1000
src_vals <- sample(n_v, n_e, replace = TRUE)
tgt_vals <- sample(n_v, n_e, replace = TRUE)

g_idx   <- GraphIdx(V = n_v, E = n_e, src = src_vals, tgt = tgt_vals)
g_noidx <- GraphNoIdx(V = n_v, E = n_e, src = src_vals, tgt = tgt_vals)
```

Compare `incident()` performance — we query every vertex:

```{r}
t_idx <- system.time(for (v in seq_len(n_v)) incident(g_idx, v, "src"))
t_noidx <- system.time(for (v in seq_len(n_v)) incident(g_noidx, v, "src"))
cat("Indexed:    ", t_idx[["elapsed"]], "s\n")
cat("No index:   ", t_noidx[["elapsed"]], "s\n")
```

The indexed version is faster because each `incident()` call is a direct hash
lookup rather than scanning all 1 000 edges.

### Index Internals

Internally, each indexed column maintains an R environment used as a hash map.
Keys are stringified values; values are integer vectors of part IDs.
The index is updated incrementally on every `set_subpart()` and `clear_subpart()`
call, so there is a small constant overhead per write.

**Rule of thumb:** index morphisms that are queried in reverse more often than
they are written.

## Query DSL Patterns

The `From() |> Where() |> Select()` pipeline supports chaining multiple
`Where()` clauses and using any binary comparison operator.

### Chained Filters

Each `Where()` narrows the result set further (logical AND):

```{r}
From(db, "Enrolment") |>
  Where("course", `==`, 2) |>
  Where("grade", `!=`, "A-") |>
  Select("student", "course", "grade")
```

### Using Different Operators

```{r}
From(fw, "Species") |>
  Where("trophic", `>=`, 2) |>
  Select("name", "trophic")
```

### Pattern Matching with grepl

`Where()` accepts any function that takes two arguments and returns a logical
vector. This means you can use `grepl` for pattern matching:

```{r}
From(db, "Student") |>
  Where("sname", grepl, "^[AB]") |>
  Select("sname")
```

### Combining incident() with the DSL

Use `incident()` to get IDs, then filter further with the DSL:

```{r}
# Edges from vertex 1 in the benchmark graph, with target > 50
v1_edges <- incident(g_idx, 1, "src")
# Build a query manually starting from those IDs
q <- structure(list(acs = g_idx, ob = "E", ids = v1_edges), class = "acset_query")
result <- Where(q, "tgt", `>`, 50) |> Select("src", "tgt")
head(result)
```

## Deletion and Cascading

### Pop-and-Swap in Detail

`rem_part()` uses a **pop-and-swap** strategy: the last part in the object
takes the deleted part's ID. This is O(1) but **reorders IDs**.

```{r}
g <- GraphIdx(V = 5, E = 4,
              src = c(1, 2, 3, 4),
              tgt = c(2, 3, 4, 5))
cat("Before deletion:\n")
as_data_frame(g, "E")
```

Remove vertex 2. Vertex 5 (the last) takes ID 2. Edges that pointed *to*
vertex 2 are cleared (`NA`); edges that pointed to vertex 5 are remapped
to 2:

```{r}
rem_part(g, "V", 2)
cat("Vertices after removing V#2:\n")
as_data_frame(g, "V")
cat("\nEdges after removing V#2:\n")
as_data_frame(g, "E")
```

Notice edge 1 (`src=1, tgt=2`) now has `tgt=NA` because its target was
deleted, while edge 4 (`src=4, tgt=5`) now shows `tgt=2` because vertex 5
was renumbered to 2.

### Multi-Level Cascading

Cascading deletion propagates recursively. Consider a schema with three levels:

```{r}
SchThreeLevel <- BasicSchema(
  obs       = c("A", "B", "C"),
  homs      = list(hom("ab", "B", "A"),
                   hom("bc", "C", "B")),
  attrtypes = c("String"),
  attrs     = list(attr_spec("label", "A", "String"))
)
ThreeLevel <- acset_type(SchThreeLevel, index = c("ab", "bc"))

tl <- ThreeLevel(
  A = 2, B = 3, C = 4,
  label = c("root1", "root2"),
  ab    = c(1, 1, 2),
  bc    = c(1, 2, 3)
)
cat("Before: A =", nparts(tl, "A"),
    ", B =", nparts(tl, "B"),
    ", C =", nparts(tl, "C"), "\n")
```

Removing A#1 cascades: first its dependent B parts are cascade-deleted (which
in turn cascade-deletes their dependent C parts), then A#1 itself is removed.

```{r}
cascading_rem_part(tl, "A", 1)
cat("After:  A =", nparts(tl, "A"),
    ", B =", nparts(tl, "B"),
    ", C =", nparts(tl, "C"), "\n")
tables(tl)
```

### Non-Cascading Deletion Leaves NAs

When you use `rem_part()` instead of `cascading_rem_part()`, dependent
morphisms are set to `NA` rather than having their parts deleted. This can
leave orphaned references:

```{r}
g2 <- GraphIdx(V = 3, E = 2, src = c(1, 2), tgt = c(2, 3))
rem_part(g2, "V", 2)
as_data_frame(g2, "E")
```

The `NA` values indicate broken foreign keys. Use `cascading_rem_part()` when
you want referential integrity maintained automatically.

## JSON Serialization Round-Trips

### Data Format

`generate_json_acset()` produces a nested list that maps directly to JSON.
Each object becomes an array of records with an `_id` field:

```{r}
small <- FSM(
  State      = 2,
  Transition = 1,
  state_name = c("On", "Off"),
  source     = 1,
  target     = 2,
  event      = "toggle"
)
json_data <- generate_json_acset(small)
str(json_data, max.level = 3)
```

### Full Round-Trip

```{r}
json_str <- jsonlite::toJSON(json_data, auto_unbox = TRUE, pretty = TRUE)
cat(json_str)
```

Parse it back:

```{r}
parsed_data <- jsonlite::fromJSON(json_str, simplifyVector = FALSE)
small2 <- parse_json_acset(FSM, parsed_data)
acset_equal(small, small2)
```

### File-Based Round-Trip

```{r}
tmp <- tempfile(fileext = ".json")
write_json_acset(small, tmp)
small3 <- read_json_acset(FSM, tmp)
acset_equal(small, small3)
```

```{r include = FALSE}
unlink(tmp)
```

### Schema Serialization

The schema itself can be serialized independently, which is useful for
transmitting the schema definition to other ACSet implementations
(Julia's ACSets.jl, Python's py-acsets):

```{r}
schema_json <- generate_json_schema(SchFSM)
str(schema_json, max.level = 2)
```

Round-trip the schema:

```{r}
schema_rt <- parse_json_schema(schema_json)
identical(objects(SchFSM), objects(schema_rt))
identical(homs(SchFSM), homs(schema_rt))
```

### Cross-Language Interoperability

The JSON format is compatible with:

| Language   | Package         |
|------------|-----------------|
| Julia      | ACSets.jl       |
| Python     | py-acsets        |
| TypeScript | ts-acsets        |
| Java       | acsets4j         |

Each object is serialized as an array of `{_id, ...}` records. Morphism values
are integer part IDs; attributes are stored as their natural JSON types. This
means you can `write_json_acset()` in R and read the file in Julia or Python
with the matching schema.

## Composition Patterns

### Building Models from Components

`disjoint_union()` combines two ACSets with the same schema. IDs from the
second ACSet are offset so there are no collisions. This lets you build
complex models from reusable pieces.

```{r}
# Two small food chains
chain1 <- FoodWeb(
  Species = 2, Predation = 1,
  name = c("Algae", "Shrimp"), trophic = c(1, 2),
  predator = 2, prey = 1, rate = 0.4
)
chain2 <- FoodWeb(
  Species = 2, Predation = 1,
  name = c("Plankton", "Fish"), trophic = c(1, 2),
  predator = 2, prey = 1, rate = 0.6
)

combined <- disjoint_union(chain1, chain2)
tables(combined)
```

After the union, species from `chain2` have IDs 3 and 4. Morphism values are
automatically remapped: the predation link `(predator=2, prey=1)` from
`chain2` becomes `(predator=4, prey=3)`.

### Copy-and-Extend

Use `copy_acset()` to create an independent snapshot, then extend it:

```{r}
extended <- copy_acset(combined)
# Add a top predator that eats both shrimp and fish
add_part(extended, "Species", name = "Shark", trophic = 3)
add_part(extended, "Predation", predator = 5, prey = 2, rate = 0.2)
add_part(extended, "Predation", predator = 5, prey = 4, rate = 0.3)
tables(extended)
```

The original `combined` is unchanged:

```{r}
nparts(combined, "Species")
nparts(extended, "Species")
```

### Iterative Assembly

You can assemble large models by iteratively unioning small components:

```{r}
SchChain <- BasicSchema(
  obs       = c("Node", "Link"),
  homs      = list(hom("from", "Link", "Node"),
                   hom("to",   "Link", "Node")),
  attrtypes = c("Numeric"),
  attrs     = list(attr_spec("value", "Node", "Numeric"))
)
Chain <- acset_type(SchChain, index = c("from", "to"))

# Build a 3-segment chain by unioning single links
segment <- function(val1, val2) {
  Chain(Node = 2, Link = 1, value = c(val1, val2), from = 1, to = 2)
}

chain <- segment(1, 2)
chain <- disjoint_union(chain, segment(3, 4))
chain <- disjoint_union(chain, segment(5, 6))
tables(chain)
```

## Schema Design Best Practices

### Attributes vs. Objects

Use **attributes** for data that doesn't need its own identity or
foreign-key relationships (labels, weights, timestamps). Use **objects** when
the data participates in relationships or needs to be referenced by other
objects.

```{r}
# Good: colour as attribute (just a label)
SchColoured <- BasicSchema(
  obs       = c("V", "E"),
  homs      = list(hom("src", "E", "V"), hom("tgt", "E", "V")),
  attrtypes = c("String"),
  attrs     = list(attr_spec("colour", "V", "String"))
)

# Better if colours are shared and queried: colour as object
SchColourObj <- BasicSchema(
  obs       = c("V", "E", "Colour"),
  homs      = list(hom("src", "E", "V"),
                   hom("tgt", "E", "V"),
                   hom("vcolour", "V", "Colour")),
  attrtypes = c("String"),
  attrs     = list(attr_spec("cname", "Colour", "String"))
)
```

When colour is an **object**, you can use `incident()` to efficiently find all
vertices of a given colour, and changing a colour name updates all references
at once:

```{r}
ColourGraph <- acset_type(SchColourObj, index = c("src", "tgt", "vcolour"))
cg <- ColourGraph(
  V = 4, E = 3, Colour = 2,
  src = c(1, 2, 3), tgt = c(2, 3, 4),
  vcolour = c(1, 1, 2, 2),
  cname = c("red", "blue")
)

# All red vertices (colour 1)
incident(cg, 1, "vcolour")
```

### Morphism Direction

Morphisms point from the **dependent** object to the **independent** one (like
foreign keys in SQL). An edge depends on its vertices, so `src: E → V`.

If you reverse the direction (vertex → edge), you lose the guarantee that each
edge has exactly one source, and you can't represent multi-edges cleanly.

### Self-Referential Morphisms

An object can have a morphism to itself. This models trees, linked lists,
or hierarchies:

```{r}
SchTree <- BasicSchema(
  obs       = c("Node"),
  homs      = list(hom("parent", "Node", "Node")),
  attrtypes = c("String"),
  attrs     = list(attr_spec("label", "Node", "String"))
)
Tree <- acset_type(SchTree, index = c("parent"))

t <- Tree()
root  <- add_part(t, "Node", label = "root")
child1 <- add_part(t, "Node", label = "child1", parent = root)
child2 <- add_part(t, "Node", label = "child2", parent = root)
leaf   <- add_part(t, "Node", label = "leaf",   parent = child1)
```

Navigate the tree:

```{r}
# Children of root
children <- incident(t, root, "parent")
subpart(t, children, "label")

# Path from leaf to root
node <- leaf
path <- character()
while (!is.na(node)) {
  path <- c(path, subpart(t, node, "label"))
  p <- subpart(t, node, "parent")
  node <- if (is.na(p)) NA else p
}
cat(paste(path, collapse = " -> "), "\n")
```

### Many-to-Many Relationships

Use a **join object** (intermediary table) to model many-to-many
relationships. The food web and university schemas above both do this:
`Predation` links Species ↔ Species, and `Enrolment` links Student ↔ Course.

This is the standard relational-database pattern, and it works identically in
the ACSet framework — with the bonus that `incident()` and `cascading_rem_part()`
understand these relationships automatically.

### Avoiding Common Pitfalls

1. **Don't duplicate morphisms as attributes.** If you have a morphism
   `src: E → V`, don't also add an attribute `src_id` on `E`. The morphism
   already stores the integer reference.

2. **Name arrows uniquely across the whole schema.** Two different morphisms
   cannot share the same name, even if they belong to different objects. Use
   prefixes if needed (e.g., `esrc` and `etgt` for edges, `asrc` for arcs).

3. **Prefer batch operations.** `add_parts(x, "V", 1000)` is much faster
   than calling `add_part()` in a loop 1 000 times because it grows the
   internal storage once instead of 1 000 times.

```{r}
BigGraph <- acset_type(SchBench)
t_batch <- system.time({
  bg <- BigGraph()
  add_parts(bg, "V", 5000)
})
t_loop <- system.time({
  bg2 <- BigGraph()
  for (i in seq_len(5000)) add_part(bg2, "V")
})
cat("Batch:  ", t_batch[["elapsed"]], "s\n")
cat("Loop:   ", t_loop[["elapsed"]], "s\n")
```
